Staged Compilation With Two-Level Type Theory

### 2022-03-03 paper/paper.tex
typos

\newcommand{\whset}{\wh{\Set}}
\newcommand{\rexti}{\re_{\ext_1}^{-1}}

@@

\subsubsection{Action on $\blank\ext_0\blank$} We have preservation up to isomorphism:
\[
  \re_{\ext_0} : \re\,(\Gamma \ext_0 A) \simeq ((\gamma : \re\,\Gamma)\times \Tm_{0,j}\,(\re\,A\,\gamma))
\]
This is because substitutions targeting $\Gamma \ext_0 A$ are the same as
pairs of substitutions and terms, by the specification of $\blank\ext_0\blank$.

@@

\subsubsection{Object-level family structure} We interpret extended contexts as follows.
\begin{alignat*}{3}
  & (\Gamma \ext_0 A)\rel\,(\gamma,\,\alpha)\,(\rexti\,(\gamma',\,\alpha')) := (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \times (\emb{\alpha} = \alpha')
\end{alignat*}
Note that $\alpha : \Tm_{\mbbo j}\,(\ev\,A\,\gamma)$, so $\emb{\alpha} :
\Tm_{0,j}\,\emb{\ev\,A\,\gamma}$, but since $A\rel\,\gamma\rel :
\emb{\ev\,A\,\gamma} = \re\,A\,\gamma'$, we also have $\emb{\alpha} :
\Tm_{0,j}\,(\re\,A\,\gamma')$. Thus, the equation $\emb{\alpha} = \alpha'$ is
well-typed. For type substitution, we need
\[
  (A[\sigma])\rel\,\gamma\rel : \emb{\ev\,(A[\sigma])\,\gamma} = \re\,(A[\sigma])\,\gamma'.
\]
The goal type computes to $\emb{\ev\,A\,(\ev\,\sigma\,\gamma)} =
\re\,A\,(\re\,\sigma\,\gamma')$. This is obtained directly from
$A\rel\,(\sigma\rel\,\gamma\rel)$.  Similarly,
\begin{alignat*}{6}
  & (t[\sigma])\rel\,\gamma\rel &&:= t\rel\,(\sigma\rel\,\gamma\rel) && \p\rel\,(\gamma\rel,\,\alpha\rel) &&:= \gamma\rel \\
  & (\sigma,\,t)\rel\,\gamma\rel &&:= (\sigma\rel\,\gamma\rel,\,t\rel\,\gamma\rel)\hspace{2em}&& \q\rel\,(\gamma\rel,\,\alpha\rel) &&:= \alpha\rel
\end{alignat*}

@@

\begin{corollary}[Conservativity of 2LTT]\label{conservativity}
From the soundness and stability of staging, we get that $\emb{\blank}$ is
bijective on types and terms, i.e.\ $\Ty_{\mbbo j}\,\Gamma \simeq \Ty_{0,j}\,\emb{\Gamma}$ and
$\Tm_{\mbbo j}\,\Gamma\,A \simeq \Tm_{0,j}\,\emb{\Gamma}\,\emb{A}$.
\end{corollary}

@@

The multi-stage calculus of Kawata and Igarashi \cite{multi-stage-calculus}
supports staging operations and type dependencies. However, there are major
differences to the current work. First, ibid.\ does not have large elimination,
nor does it allow staged manipulation of type expressions. Second, it does not
support staged compilation in the sense of Definition \ref{def:staging}; rather,
it is a system for runtime code manipulation and evaluation (which we may call
staged \emph{computation}). In other words, programs may manipulate code
fragments during execution, in more liberal ways than in 2LTT, but there is no
judgment which ensures that a purely ``object-level'' program can be extracted
from a multi-stage program.

>>>>

\newcommand{\whset}{\wh{\Set}}
\newcommand{\rexti}{\re_{\ext_1}^{-1}}
\newcommand{\rextizero}{\re_{\ext_0}^{-1}}

@@

\subsubsection{Action on $\blank\ext_0\blank$} We have preservation up to isomorphism:
\[
  \re_{\ext_0} : \re\,(\Gamma \ext_0 A) \simeq ((\gamma : \re\,\Gamma)\times \Tm_{0,j}\,(\re\,A\,\gamma))
\]
This is because substitutions targeting $\Gamma \ext_0 A$ are the same as
pairs of substitutions and terms, by the specification of $\blank\!\ext_0\!\blank$.

@@

\subsubsection{Object-level family structure} We interpret extended contexts as follows.
\begin{alignat*}{3}
  & (\Gamma \ext_0 A)\rel\,(\gamma,\,\alpha)\,(\rextizero\,(\gamma',\,\alpha')) := (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \times (\emb{\alpha} = \alpha')
\end{alignat*}
Note that $\alpha : \Tm_{\mbbo j}\,(\ev\,A\,\gamma)$, so $\emb{\alpha} :
\Tm_{0,j}\,\emb{\ev\,A\,\gamma}$, but since $A\rel\,\gamma\rel :
\emb{\ev\,A\,\gamma} = \re\,A\,\gamma'$, we also have $\emb{\alpha} :
\Tm_{0,j}\,(\re\,A\,\gamma')$. Thus, the equation $\emb{\alpha} = \alpha'$ is
well-typed. For type substitution, we need
\[
  (A[\sigma])\rel\,\gamma\rel : \emb{\ev\,(A[\sigma])\,\gamma} = \re\,(A[\sigma])\,\gamma'.
\]
The goal type computes to $\emb{\ev\,A\,(\ev\,\sigma\,\gamma)} =
\re\,A\,(\re\,\sigma\,\gamma')$. This is obtained directly from
$A\rel\,(\sigma\rel\,\gamma\rel)$.  Similarly,
\begin{alignat*}{6}
  & (t[\sigma])\rel\,\gamma\rel &&:= t\rel\,(\sigma\rel\,\gamma\rel) && \p\rel\,(\gamma\rel,\,\alpha\rel) &&:= \gamma\rel \\
  & (\sigma,\,t)\rel\,\gamma\rel &&:= (\sigma\rel\,\gamma\rel,\,t\rel\,\gamma\rel)\hspace{2em}&& \q\rel\,(\gamma\rel,\,\alpha\rel) &&:= \alpha\rel
\end{alignat*}

@@

\begin{corollary}[Conservativity of 2LTT]\label{conservativity}
From the soundness and stability of staging, we get that $\emb{\blank}$ is
bijective on types and terms, hence $\Ty_{\mbbo j}\,\Gamma \simeq \Ty_{0,j}\,\emb{\Gamma}$ and
$\Tm_{\mbbo j}\,\Gamma\,A \simeq \Tm_{0,j}\,\emb{\Gamma}\,\emb{A}$.
\end{corollary}

@@

The multi-stage calculus of Kawata and Igarashi \cite{multi-stage-calculus}
supports staging operations and type dependencies. However, there are major
differences to the current work. First, ibid.\ does not have large elimination,
hence it does not support computing types by staging. Second, it does not
support staged compilation in the sense of Definition \ref{def:staging}; rather,
it is a system for runtime code manipulation and evaluation (which we may call
staged \emph{computation}). In other words, programs may manipulate code
fragments during execution, in more liberal ways than in 2LTT, but there is no
judgment which ensures that a purely ``object-level'' program can be extracted
from a multi-stage program.
